---
title: "PseudoCounts_DEG_Prep"
format: pdf
---
#Prepping Salmon Pseudo-count data for DESeq2

Use tximport to import and combine the quant.sh files created for each sample

```{r}
library("tximport")
library("txdbmaker")
library("GenomicFeatures")
library("rtracklayer")
library("AnnotationDbi")
library("biomaRt")
library("pathview")
library("gage")
library("gageData")
library('clusterProfiler')
library('enrichplot')
library('dplyr')

```

```{r}
#To use DESeq2 we need to first have a sample table that tells DESeq2 what samples we are interested in and an covariates we want to add to the GLM.
#First read in the sample table
# Load necessary libraries
library(knitr)

# Set the working directory
knitr::opts_knit$set(root.dir = "/Users/bridgetwong/RNAseq/")  # Change this path as needed

# Optional: Set other options
#options(stringsAsFactors = FALSE)

directory <- "/Users/bridgetwong/RNAseq/250514_FIg4Global_P21_Cortex_Counts/"
samples <- data.frame(
  samples = c("ctrl1", "mut1", "ctrl2", "mut2", "ctrl3", "mut3", "ctrl4", "mut4", "ctrl5", "mut5"),
  Condition = c("ctrl", "mut","ctrl", "mut","ctrl", "mut","ctrl", "mut", "ctrl", "mut"),

  stringsAsFactors = FALSE  # Prevent factors, if you prefer character vectors
)

# Write the data frame to a CSV file
write.csv(samples, "design_012424_4samples.csv", row.names = FALSE)

# Confirm creation
print("design_012424_4samples.csv has been created.")
samples <- read.table("design_012424_4samples.csv", sep=',', header=TRUE)
```

```{r}
#Take a look at the samples table
samples
```

Check to make sure everything is there...

```{r}
# Base directory where the quant.sf files are located
base_dir <- "/Users/bridgetwong/RNAseq/250514_FIg4Global_P21_Cortex_Counts"

# Constructing file paths for quant.sf files
files <- file.path(base_dir, samples$sample, "quant.sf")

# Print the constructed file paths to check
print(files)
#files <- file.path("/Users/bridgetwong/RNAseq/Salmon_Quants/", samples$sample, "quant.sf")
all(file.exists(files))
```

Now we need to make the tx2gene file so that the transcripts can be associated with gene IDs for gene-level summarization. For this, we need to create a transcript to gene file that associates each transcript with its gene

```{r}
library(GenomicFeatures)
library(AnnotationDbi)
library(biomaRt)
library(writexl)

# Set the GTF path
gtfPath <- file.path("/Users/bridgetwong/RNAseq", "gencode.vM10.annotation.gtf")

# Create TxDb from GTF
txdb <- makeTxDbFromGFF(file = gtfPath)

# Retrieve transcript names
k <- keys(txdb, keytype = "TXNAME")

# Map transcripts to gene IDs
tx2gene <- AnnotationDbi::select(txdb, keys = k, columns = "GENEID", keytype = "TXNAME")
head(tx2gene)

# Load biomaRt
library(biomaRt)

# Initialize BioMart for the mouse dataset (change to hsapiens_gene_ensembl for human data)
ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")

# Connect to Ensembl, using a mirror site if Ensembl is down
#ensembl<- useEnsembl(biomart = "genes", dataset = "mmusculus_gene_ensembl", mirror = "useast") # Or use 'useast'/'asia'

# Strip version numbers from Ensembl IDs in tx2gene if necessary
tx2gene$GENEID <- sub("\\..*", "", tx2gene$GENEID)

# Run BioMart query again
gene_symbols <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
                      filters = "ensembl_gene_id",
                      values = unique(tx2gene$GENEID),
                      mart = ensembl)


# Merge Gene Symbols with tx2gene and handle missing values
tx2gene <- merge(tx2gene, gene_symbols, by.x = "GENEID", by.y = "ensembl_gene_id", all.x = TRUE)
tx2gene <- tx2gene %>% 
  rename(GeneSymbol = external_gene_name, Transcript = TXNAME) %>%
  mutate(GeneSymbol = ifelse(is.na(GeneSymbol), GENEID, GeneSymbol))


```


```{r}
# Step 3: Import with tximport to include Counts, Abundance, and Length
names(files) <- paste0(samples$sample)
txi.salmon <- tximport(files, type = "salmon", tx2gene = tx2gene[, c("Transcript", "GENEID")])

# Step 4: Convert matrices to data frames and add Gene IDs, Gene Symbols, and Transcripts as columns
counts_df <- as.data.frame(txi.salmon$counts)
abundance_df <- as.data.frame(txi.salmon$abundance)
length_df <- as.data.frame(txi.salmon$length)

# Merge to add GENEID, GeneSymbol, and Transcript columns to each matrix
counts_df <- tibble::rownames_to_column(counts_df, var = "GENEID")
abundance_df <- tibble::rownames_to_column(abundance_df, var = "GENEID")
length_df <- tibble::rownames_to_column(length_df, var = "GENEID")

# Join with tx2gene to include GeneSymbol and Transcript columns
counts_df <- merge(counts_df, tx2gene, by = "GENEID", all.x = TRUE)
abundance_df <- merge(abundance_df, tx2gene, by = "GENEID", all.x = TRUE)
length_df <- merge(length_df, tx2gene, by = "GENEID", all.x = TRUE)

# Step 5: Remove duplicate gene names (keep first occurrence)
counts_df <- counts_df %>% distinct(GeneSymbol, .keep_all = TRUE)
abundance_df <- abundance_df %>% distinct(GeneSymbol, .keep_all = TRUE)
length_df <- length_df %>% distinct(GeneSymbol, .keep_all = TRUE)


# Step 6: Export to Excel with each matrix as a separate sheet
txi_data <- list(
  Counts = counts_df,
  Abundance = abundance_df,
  Length = length_df
)

```
```{r}
#Check that genes are annotated
print(dim(txi.salmon$counts))

```

```{r}

# Step 1: Define refined patterns for unwanted genes
genes_to_remove <- c("^mt", "^Rpl", "^Rps", "^GM", "^Gm", "Rik$")

# Step 2: Trim whitespace from GeneSymbol column
counts_df$GeneSymbol <- trimws(counts_df$GeneSymbol)
abundance_df$GeneSymbol <- trimws(abundance_df$GeneSymbol)
length_df$GeneSymbol <- trimws(length_df$GeneSymbol)

# Step 3: Filter out unwanted genes with case-insensitivity
counts_df <- counts_df[!grepl(paste(genes_to_remove, collapse = "|"), counts_df$GeneSymbol, ignore.case = TRUE), ]
abundance_df <- abundance_df[!grepl(paste(genes_to_remove, collapse = "|"), abundance_df$GeneSymbol, ignore.case = TRUE), ]
length_df <- length_df[!grepl(paste(genes_to_remove, collapse = "|"), length_df$GeneSymbol, ignore.case = TRUE), ]

# Step 4: Verify that unwanted genes are removed
sum(grepl("^Gm", counts_df$GeneSymbol, ignore.case = TRUE))  # Should be 0
sum(grepl("Rik$", counts_df$GeneSymbol, ignore.case = TRUE))  # Should be 0

# Step 5: Export to Excel
txi_data_filtered <- list(
  Counts = counts_df,
  Abundance = abundance_df,
  Length = length_df
)


# Check the first few rows of the filtered data to ensure the unwanted genes are removed
head(counts_df)
head(abundance_df)
head(length_df)


# Step 1: Reorder the columns to move GeneID to the last column
counts_df <- counts_df[, c(setdiff(names(counts_df), "GENEID"), "GENEID")]
abundance_df <- abundance_df[, c(setdiff(names(abundance_df), "GENEID"), "GENEID")]
length_df <- length_df[, c(setdiff(names(length_df), "GENEID"), "GENEID")]

# Step 2: Check if GeneID is moved to the last column
head(counts_df)
head(abundance_df)
head(length_df)

# Step 3: Export the updated data to Excel
txi_data_reordered <- list(
  Counts = counts_df,
  Abundance = abundance_df,
  Length = length_df
)


```



```{r}
# Check the first few row names and summary of counts
head(rownames(txi.salmon$counts))  # Ensure these are gene symbols
summary(txi.salmon$counts)        # Ensure there are non-zero counts

```
```{r}
# Ensure tx2gene has unique mappings between Ensembl IDs and Gene Symbols
rownames_mapping <- tx2gene$GeneSymbol
names(rownames_mapping) <- tx2gene$GENEID

```

```{r}
# Set row names to gene symbols for each matrix in txi.salmon
rownames(txi.salmon$counts) <- rownames_mapping[rownames(txi.salmon$counts)]
rownames(txi.salmon$abundance) <- rownames_mapping[rownames(txi.salmon$abundance)]
rownames(txi.salmon$length) <- rownames_mapping[rownames(txi.salmon$length)]

```

```{r}
# Print first few rows to confirm
head(rownames(txi.salmon$counts))
head(rownames(txi.salmon$abundance))
head(rownames(txi.salmon$length))

# Step 1: Define unwanted gene patterns
genes_to_remove <- c("^mt", "^Rpl", "^Rps", "^GM", "^Gm", "Rik$")

# Step 2: Filter counts, abundance, and length in txi.salmon
filtered_counts <- txi.salmon$counts[!grepl(paste(genes_to_remove, collapse = "|"), rownames(txi.salmon$counts), ignore.case = TRUE), ]
filtered_abundance <- txi.salmon$abundance[!grepl(paste(genes_to_remove, collapse = "|"), rownames(txi.salmon$abundance), ignore.case = TRUE), ]
filtered_length <- txi.salmon$length[!grepl(paste(genes_to_remove, collapse = "|"), rownames(txi.salmon$length), ignore.case = TRUE), ]

# Step 3: Update txi.salmon with filtered data
txi.salmon$counts <- filtered_counts
txi.salmon$abundance <- filtered_abundance
txi.salmon$length <- filtered_length

# Verify filtering
print(dim(txi.salmon$counts))  # Check the dimensions after filtering

```

##DESEQ2

```{r}
library("DESeq2")
library("ggplot2")
library("pheatmap")
library("RColorBrewer")
library("PoiClaClu")
library("genefilter")
library("EnhancedVolcano")
library("openxlsx")
library("ggrepel")
```

```{r}
# samples <- read.table("design_012424_4samples.csv", sep=',', header=TRUE)
# print(head(samples))  # Display the first few rows
```

```{r}
samples$Condition <- factor(samples$Condition)
# Check if sample identifiers match
setdiff(rownames(samples), colnames(txi.salmon$counts))  # Should return an empty vector
setdiff(colnames(txi.salmon$counts), rownames(samples))  # Should return an empty vector

```

```{r}
# Set row names of samples to match the column names of txi.salmon$counts
rownames(samples) <- colnames(txi.salmon$counts)

# Now check if the row names of samples and column names of counts match
setdiff(rownames(samples), colnames(txi.salmon$counts))  # Should return an empty vector
setdiff(colnames(txi.salmon$counts), rownames(samples))  # Should return an empty vector

any(is.na(rownames(samples)))  # Should return FALSE if there are no missing row names
# Check that the conditions in 'samples' match the column names in 'txi.salmon$counts'
all(rownames(samples) == colnames(txi.salmon$counts))  # Should return TRUE

# Make sure that the row names of 'samples' match column names of 'txi.salmon$counts'
identical(rownames(samples), colnames(txi.salmon$counts))  # Should return TRUE
# Check for the Condition column
head(samples$Condition)

# Make sure 'Condition' is a factor
samples$Condition <- factor(samples$Condition)

# Check for any NA values in rownames of the counts data
any(is.na(rownames(txi.salmon$counts)))  # Should return FALSE

# Check for any NA values in rownames of the samples metadata
any(is.na(rownames(samples)))  # Should return FALSE

na_rows <- which(is.na(rownames(txi.salmon$counts)))
na_rows

print(dim(txi.salmon$counts))

```


```{r}

# Check that the design formula is correct
DESeq2CDS <- DESeqDataSetFromTximport(
  txi = txi.salmon,  # txi.salmon contains the count data
  colData = samples,  # sample metadata
  design = ~ Condition  # design formula to specify the condition variable
)
dds <- DESeq(DESeq2CDS)


#dds <- DESeqDataSetFromMatrix(countData = countData, colData = samples, design = ~ Condition)

```

```{r}
#PCA Plots
vsd <- vst(dds, blind=FALSE)
pcaData <- plotPCA(vsd, intgroup = c("Condition"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=Condition)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```

```{r}
##Some QC comparisons of our data, all plots should be in a  similar positive linear direction. 
par( mfrow = c( 1, 1 ) )
dds <- estimateSizeFactors(dds)
plot(log2(counts(dds, normalized=TRUE)[,1:2] + 1), pch=16, cex=0.3)
plot(log2(counts(dds, normalized=TRUE)[,3:4] + 1), pch=16, cex=0.3)

vst <- vst(dds, blind=FALSE)
plot(assay(vst)[,1:2], pch=16, cex=0.3)
sampleDists <- dist( t( assay(vst) ) )
sampleDists
```

```{r}
##Heatmap visualization
sampleDistMatrix <- as.matrix( sampleDists )
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

```{r}
sampleNames =c( 'ctrl1',  'ctrl2',  "ctrl3",  "ctrl4",  "ctrl5",'mut1','mut2',"mut3","mut4", "mut5")
```

```{r}
## PoiClaClu
poisd <- PoissonDistance(t(counts(dds)))
samplePoisDistMatrix <- as.matrix( poisd$dd )
rownames(samplePoisDistMatrix) <- paste( vst$dex, vst$cell, sep="-" )
colnames(samplePoisDistMatrix) <- sampleNames
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows=poisd$dd,
         clustering_distance_cols=poisd$dd,
         col=colors)
```
```{r}
plotPCA(vst,  intgroup = c("Condition"))
(data <- plotPCA(vst, intgroup = c( "Condition"), returnData=TRUE))
percentVar <- round(100 * attr(data, "percentVar"))
mdsPoisData <- data.frame(cmdscale(samplePoisDistMatrix))
mdsPois <- cbind(mdsPoisData, as.data.frame(colData(dds)))
ggplot(mdsPois, aes(X1,X2,shape=Condition)) + geom_point(size=3)
```

```{r}
## Further Visualizations
(res <- results(dds))
mcols(res, use.names=TRUE)
summary(res)
res.05 <- results(dds, alpha=.05)
table(res.05$padj < .05)
resLFC1 <- results(dds, lfcThreshold=1)
topGene <- rownames(res)[which.min(res$padj)]
print(topGene)
plotCounts(dds, gene=topGene, intgroup=c("Condition"))

data <- plotCounts(dds, gene=topGene, intgroup=c("Condition"), returnData=TRUE)
ggplot(data, aes(x=Condition, y=count)) +
  scale_y_log10() +
  geom_point(position=position_jitter(width=.1,height=0), size=3)

ggplot(data, aes(x=Condition, y=count, fill=Condition)) +
  scale_y_log10() +
  geom_dotplot(binaxis="y", stackdir="center", binwidth = 1/30)

ggplot(data, aes(x=Condition, y=count, color=Condition, group=Condition)) +
  scale_y_log10() + geom_point(size=3) + geom_line()
```

```{r}
#Mean-average plot, genes with low read counts show higher variability than highly expressed genes, resulting in a strong log2FC even though are likely not really differentially expressed

plotMA(res, ylim=c(-5,5))

topGene <- rownames(resLFC1)[which.min(resLFC1$padj)]
with(resLFC1[topGene, ], {
  points(baseMean, log2FoldChange, col="dodgerblue", cex=2, lwd=2)
  text(baseMean, log2FoldChange, topGene, pos=2, col="dodgerblue")
})

#Rough idea of proportion of genes that are differentially expressed
hist(res$pvalue[res$baseMean > 1], breaks=0:20/20, col="grey50", border="white")
```

```{r}
topVarGenes <- head(order(rowVars(assay(vst)),decreasing=TRUE),40)
mat <- assay(vst)[ topVarGenes, ]
mat <- mat - rowMeans(mat)
df <- as.data.frame(colData(vst)[,c("Condition","Condition")])
pheatmap(mat, annotation_col=df, fontsize_row=5)
```

```{r}
qs <- c(0, quantile(resLFC1$baseMean[resLFC1$baseMean > 0], 0:6/6))
bins <- cut(resLFC1$baseMean, qs) 
levels(bins) <- paste0("~",round(signif(.5*qs[-1] + .5*qs[-length(qs)],2)))
ratios <- tapply(resLFC1$pvalue, bins, function(p) mean(p < .05, na.rm=TRUE))
barplot(ratios, xlab="mean normalized count", ylab="ratio of small p values")
plotDispEsts(dds, ylim = c(1e-6, 1e1))
```

```{r}
hist(res$pvalue, breaks = 20, col = "grey")
qs <- c( 0, quantile( res$baseMean[res$baseMean > 0], 0:7/7 ) )
bins <- cut( res$baseMean, qs )
levels(bins) <- paste0("~",round(.5*qs[-1] + .5*qs[-length(qs)]))
ratios <- tapply( res$pvalue, bins, function(p) mean( p < .01, na.rm=TRUE ) )
barplot(ratios, xlab="mean normalized count", ylab="ratio of small p values")
```

```{r}
path_out = "/Users/bridgetwong/RNAseq"
Main_Output = "/Users/bridgetwong/RNAseq"
outputPrefix <- "DESeq2"
```


```{r}

setwd(path_out)
## Writing Outputs
res[1:2,]
write.csv(as.data.frame(res), file="results.csv" )
res <- results(dds)
res1= subset(res, padj<1)
resdata <- merge(as.data.frame(res1), as.data.frame(counts(dds,normalized =TRUE)), by = 'row.names', sort = FALSE)
names(resdata)[1] <- 'ensembl_gene_id'

write.csv(resdata,  file = paste0(outputPrefix, "-results-with-normalized.csv"))
write.table(as.data.frame(counts(dds),normalized=T), file = paste0(outputPrefix, "_normalized_counts.txt"), sep = '\t')
pcaData <- plotPCA(vst, intgroup=c("Condition"), returnData=TRUE)
write.table(as.data.frame(pcaData,normalized=T), file = paste0(outputPrefix, "_PCArawData.txt"), sep = '\t')

setwd(Main_Output)
```

```{r fig.height=8}
#can change 20 to 15, 10, 30 etc. 
top20 <- head(resdata[order(resdata$padj),], 1)
Labels20 <- top20$ensembl_gene_id
EnhancedVolcano(resdata,
                lab = resdata$ensembl_gene_id,
                x = 'log2FoldChange',
                y = 'padj',
                title = 'WT vs KO',
                pCutoff = 0.05,
                FCcutoff = 0.3,
                selectLab = c(Labels20),
                pointSize = 1.0,
                labSize = 2.5,
                boxedLabels = TRUE,
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 1, 
                drawConnectors = TRUE,
                widthConnectors = 0.2,
                colConnectors = 'black')
```

```{r}
## Main Excel Output of DEGs.  
rownames(resdata) <- NULL
wb <- createWorkbook("250514_FIg4Global_P21_Cortex_Salmon_Results.xlsx")
addWorksheet(wb, "Output", gridLines = TRUE) 
writeData(wb, sheet = "Output", resdata, rowNames = FALSE)
setColWidths(wb, sheet = "Output", cols = 1:12, widths = "15")
saveWorkbook(wb, "250514_FIg4Global_P21_Cortex_Salmon_Results.xlsx", overwrite = TRUE)

padj <- subset(resdata, padj<0.05)
addWorksheet(wb, "padj<0.05", gridLines = TRUE) 
writeData(wb, sheet = "padj<0.05", padj, rowNames = FALSE)
setColWidths(wb, sheet = "padj<0.05", cols = 1:12, widths = "15")
saveWorkbook(wb, "250514_FIg4Global_P21_Cortex_Salmon_Results.xlsx", overwrite = TRUE)

log2fc <- subset(padj, log2FoldChange>0.3 | log2FoldChange<(-0.3))
addWorksheet(wb, "log2fc_and_padj", gridLines = TRUE) 
writeData(wb, sheet = "log2fc_and_padj", log2fc, rowNames = FALSE)
setColWidths(wb, sheet = "log2fc_and_padj", cols = 1:12, widths = "15")
saveWorkbook(wb, "250514_FIg4Global_P21_Cortex_Salmon_Results.xlsx", overwrite = TRUE)

UP <- subset(log2fc, log2FoldChange>0.3)
addWorksheet(wb, "Upregulated", gridLines = TRUE) 
writeData(wb, sheet = "Upregulated", UP, rowNames = FALSE)
setColWidths(wb, sheet = "Upregulated", cols = 1:12, widths = "15")
saveWorkbook(wb, "250514_FIg4Global_P21_Cortex_Salmon_Results.xlsx", overwrite = TRUE)

DOWN <- subset(log2fc, log2FoldChange< -0.3)
addWorksheet(wb, "Downregulated", gridLines = TRUE) 
writeData(wb, sheet = "Downregulated", DOWN, rowNames = FALSE)
setColWidths(wb, sheet = "Downregulated", cols = 1:12, widths = "15")
saveWorkbook(wb, "250514_FIg4Global_P21_Cortex_Salmon_Results.xlsx", overwrite = TRUE)
```

```{r}
# Create background dataset for hypergeometric testing using all genes tested for significance in the results
all_genes <- as.character(rownames(res1))
# Extract significant results
signif_res <- res[res$padj < 0.05 & !is.na(res$padj), ]
signif_genes <- as.character(rownames(signif_res))
```

```{r}
# Run GO enrichment analysis


library(org.Mm.eg.db)
ego <- enrichGO(gene = signif_genes,
universe = all_genes,
keyType = "SYMBOL",
OrgDb = org.Mm.eg.db,
ont = "all",
pAdjustMethod = "BH",
qvalueCutoff = 0.05,
readable = TRUE)

# Output results from GO analysis to a table
cluster_summary <- data.frame(ego)

```

```{r, fig.width=10, fig.height=15, dpi=300}


dotplot(ego, showCategory=10)


#Enrichment GO plot
x2 <- pairwise_termsim(ego) 
emapplot(x2, showCategory = 5)


# To color genes by log2 fold changes
cnetplot(ego, categorySize="pvalue", showCategory = 2, vertex.label.font=6)

```

Now to separate DEGs into up and down for Go analysis
```{r}
resSig <- res[which(res$padj < 0.05), ]
# Separate upregulated and downregulated genes
upregulated_genes <- rownames(resSig[which(resSig$log2FoldChange > 0), ])
downregulated_genes <- rownames(resSig[which(resSig$log2FoldChange < 0), ])

# GO enrichment analysis for upregulated genes
ego_up <- enrichGO(gene = upregulated_genes,
                   OrgDb = org.Mm.eg.db,
                   universe = all_genes,
                   keyType = "SYMBOL",
                   ont = "ALL",
                   pAdjustMethod = "BH",
                   qvalueCutoff = 0.05)

# GO enrichment analysis for downregulated genes
ego_down <- enrichGO(gene = downregulated_genes,
                     OrgDb = org.Mm.eg.db,
                     universe = all_genes,
                     keyType = "SYMBOL",
                     ont = "ALL",
                     pAdjustMethod = "BH",
                     qvalueCutoff = 0.05)
```

```{r, fig.width=10, fig.height=15, dpi=300}
#Visualizations
dotplot(ego_up, showCategory = 20, title = "GO Enrichment for Upregulated Genes")
dotplot(ego_down, showCategory = 20, title = "GO Enrichment for Downregulated Genes")
```
```{r}
# Create background dataset for hypergeometric testing using all genes tested for significance in the results
all_genes <- as.character(rownames(res1))
# Extract significant results
signif_res <- res[res$padj < 0.05 & !is.na(res$padj), ]
signif_genes <- as.character(rownames(signif_res))
```

```{r}
# Run GO enrichment analysis
ego <- enrichGO(gene = signif_genes,
universe = all_genes,
keyType = "SYMBOL",
OrgDb = org.Mm.eg.db,
ont = "CC",
pAdjustMethod = "BH",
qvalueCutoff = 0.05,
readable = TRUE)
# Output results from GO analysis to a table
cluster_summary <- data.frame(ego)
```

```{r, fig.width=10, fig.height=15, dpi=300}
#Visualizations
dotplot(ego, showCategory=20)
```

Now to separate DEGs into up and down for Go analysis
```{r}
resSig <- res[which(res$padj < 0.05), ]
# Separate upregulated and downregulated genes
upregulated_genes <- rownames(resSig[which(resSig$log2FoldChange > 0), ])
downregulated_genes <- rownames(resSig[which(resSig$log2FoldChange < 0), ])
```


```{r}
# GO enrichment analysis for upregulated genes
ego_up <- enrichGO(gene = upregulated_genes,
                   OrgDb = org.Mm.eg.db,
                   universe = all_genes,
                   keyType = "SYMBOL",
                   ont = "BP",
                   pAdjustMethod = "BH",
                   qvalueCutoff = 0.05)

# GO enrichment analysis for downregulated genes
ego_down <- enrichGO(gene = downregulated_genes,
                     OrgDb = org.Mm.eg.db,
                     universe = all_genes,
                     keyType = "SYMBOL",
                     ont = "BP",
                     pAdjustMethod = "BH",
                     qvalueCutoff = 0.05)
```

```{r, fig.width=10, fig.height=15, dpi=300}

library(clusterProfiler)
library(org.Mm.eg.db)
library(GOSemSim)
library(pheatmap)
library(rrvgo)
library(clusterProfiler)
library(org.Mm.eg.db)  # For mouse data, change to your organism


# ---- 1) GO Enrichment ----
ego_up <- enrichGO(gene = upregulated_genes,
                   OrgDb = org.Mm.eg.db,
                   universe = all_genes,
                   keyType = "SYMBOL",
                   ont = "BP",
                   pAdjustMethod = "BH",
                   qvalueCutoff = 0.05)

ego_down <- enrichGO(gene = downregulated_genes,
                     OrgDb = org.Mm.eg.db,
                     universe = all_genes,
                     keyType = "SYMBOL",
                     ont = "BP",
                     pAdjustMethod = "BH",
                     qvalueCutoff = 0.05)


```

```{r}

#Calculate a similarity matrix between terms
simMatrixup <- calculateSimMatrix(ego_up$ID,
                                orgdb=org.Mm.eg.db,
                                keytype = "ENTREZID",
                                ont="BP",
                                method="Rel")

simMatrixdown <- calculateSimMatrix(ego_down$ID,
                                orgdb=org.Mm.eg.db,
                                keytype = "ENTREZID",
                                ont="BP",
                                method="Rel")

scoresup <- setNames(-log10(ego_up$qvalue), ego_up$ID)
reducedTermsup <- reduceSimMatrix(simMatrixup,
                                scoresup,
                                threshold=0.7,
                                orgdb="org.Mm.eg.db")

scoresdown <- setNames(-log10(ego_down$qvalue), ego_down$ID)
reducedTermsdown <- reduceSimMatrix(simMatrixdown,
                                scoresdown,
                                threshold=0.7,
                                orgdb="org.Mm.eg.db")

```

```{r, fig.width=10, fig.height=15, dpi=300}
heatmapPlot(simMatrixup,
            reducedTermsup,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=8)
heatmapPlot(simMatrixdown,
            reducedTermsdown,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=8)

```

```{r, fig.width=10, fig.height=15, dpi=300}
scatterPlot(simMatrixup, reducedTermsup, size = "score")
scatterPlot(simMatrixdown, reducedTermsdown, size = "score")
```


```{r, fig.width=10, fig.height=15, dpi=300}
# Load required packages
treemapPlot(reducedTermsup)
treemapPlot(reducedTermsdown)
```

```{r, fig.width=10, fig.height=15, dpi=300}

library(clusterProfiler)
library(org.Mm.eg.db)
library(GOSemSim)
library(pheatmap)
library(rrvgo)
library(clusterProfiler)
library(org.Mm.eg.db)  # For mouse data, change to your organism


# ---- 1) GO Enrichment ----
ego_up <- enrichGO(gene = upregulated_genes,
                   OrgDb = org.Mm.eg.db,
                   universe = all_genes,
                   keyType = "SYMBOL",
                   ont = "CC",
                   pAdjustMethod = "BH",
                   qvalueCutoff = 0.05)

ego_down <- enrichGO(gene = downregulated_genes,
                     OrgDb = org.Mm.eg.db,
                     universe = all_genes,
                     keyType = "SYMBOL",
                     ont = "CC",
                     pAdjustMethod = "BH",
                     qvalueCutoff = 0.05)


```

```{r}

#Calculate a similarity matrix between terms
simMatrixup <- calculateSimMatrix(ego_up$ID,
                                orgdb=org.Mm.eg.db,
                                keytype = "ENTREZID",
                                ont="CC",
                                method="Rel")

simMatrixdown <- calculateSimMatrix(ego_down$ID,
                                orgdb=org.Mm.eg.db,
                                keytype = "ENTREZID",
                                ont="CC",
                                method="Rel")

scoresup <- setNames(-log10(ego_up$qvalue), ego_up$ID)
reducedTermsup <- reduceSimMatrix(simMatrixup,
                                scoresup,
                                threshold=0.7,
                                orgdb="org.Mm.eg.db")

scoresdown <- setNames(-log10(ego_down$qvalue), ego_down$ID)
reducedTermsdown <- reduceSimMatrix(simMatrixdown,
                                scoresdown,
                                threshold=0.7,
                                orgdb="org.Mm.eg.db")

```

```{r, fig.width=10, fig.height=15, dpi=300}
heatmapPlot(simMatrixup,
            reducedTermsup,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=8)
heatmapPlot(simMatrixdown,
            reducedTermsdown,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=8)

```

```{r, fig.width=10, fig.height=15, dpi=300}
scatterPlot(simMatrixup, reducedTermsup, size = "score")
scatterPlot(simMatrixdown, reducedTermsdown, size = "score")
```


```{r, fig.width=10, fig.height=15, dpi=300}
# Load required packages
treemapPlot(reducedTermsup)
treemapPlot(reducedTermsdown)
```

```{r, fig.width=10, fig.height=15, dpi=300}

library(clusterProfiler)
library(org.Mm.eg.db)
library(GOSemSim)
library(pheatmap)
library(rrvgo)
library(clusterProfiler)
library(org.Mm.eg.db)  # For mouse data, change to your organism


# ---- 1) GO Enrichment ----
ego_up <- enrichGO(gene = upregulated_genes,
                   OrgDb = org.Mm.eg.db,
                   universe = all_genes,
                   keyType = "SYMBOL",
                   ont = "MF",
                   pAdjustMethod = "BH",
                   qvalueCutoff = 0.05)

ego_down <- enrichGO(gene = downregulated_genes,
                     OrgDb = org.Mm.eg.db,
                     universe = all_genes,
                     keyType = "SYMBOL",
                     ont = "MF",
                     pAdjustMethod = "BH",
                     qvalueCutoff = 0.05)


```

```{r}

#Calculate a similarity matrix between terms
simMatrixup <- calculateSimMatrix(ego_up$ID,
                                orgdb=org.Mm.eg.db,
                                keytype = "ENTREZID",
                                ont="MF",
                                method="Rel")

simMatrixdown <- calculateSimMatrix(ego_down$ID,
                                orgdb=org.Mm.eg.db,
                                keytype = "ENTREZID",
                                ont="MF",
                                method="Rel")

scoresup <- setNames(-log10(ego_up$qvalue), ego_up$ID)
reducedTermsup <- reduceSimMatrix(simMatrixup,
                                scoresup,
                                threshold=0.7,
                                orgdb="org.Mm.eg.db")

scoresdown <- setNames(-log10(ego_down$qvalue), ego_down$ID)
reducedTermsdown <- reduceSimMatrix(simMatrixdown,
                                scoresdown,
                                threshold=0.7,
                                orgdb="org.Mm.eg.db")

```

```{r, fig.width=10, fig.height=15, dpi=300}
heatmapPlot(simMatrixup,
            reducedTermsup,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=8)
heatmapPlot(simMatrixdown,
            reducedTermsdown,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=8)

```

```{r, fig.width=10, fig.height=15, dpi=300}
scatterPlot(simMatrixup, reducedTermsup, size = "score")
scatterPlot(simMatrixdown, reducedTermsdown, size = "score")
```


```{r, fig.width=10, fig.height=15, dpi=300}
# Load required packages
treemapPlot(reducedTermsup)
treemapPlot(reducedTermsdown)
```